---
title: "Competition"
description: |
  Competition with Lotka-Volterra models
author: Nathan Jud
date: 2024-05-13
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

### Learning Objectives:
1. Understand the basics of constructing a phase plane diagram for a Lotka-Volterra style model.
2. Differentiate between intraspecific and interspecific competition in ecological systems.
3. Identify zero net growth isoclines and their significance in population dynamics.
4. Explore the concept of niche and its relationship with the assumptions of the Lotka-Volterra model.
5. Introduce multi-species coexistence models as an extension of Lotka-Volterra.

### Preparation Questions:
1. What are the key differences between intraspecific and interspecific competition?
2. How do zero net growth isoclines help us understand the stability of population dynamics?
3. In what ways do the assumptions of the Lotka-Volterra model simplify real-world ecological systems?
4. Can you think of examples where species coexist in an ecosystem despite competing for the same resources?
5. How might adding more species to the Lotka-Volterra model affect its complexity and predictive power?

```{r echo=FALSE}
# Load required libraries
library(ggplot2)
```

```{r}
# Define Lotka-Volterra equations
LV_equations <- function(t, N, params) {
  with(as.list(c(N, params)), {
    dN1 <- r1 * N1 * (1 - (N1 * alpha1 + N2 * beta1))
    dN2 <- r2 * N2 * (1 - (N1 * alpha2 + N2 * beta2))
    return(list(c(dN1, dN2)))
  })
}
```

```{r}
# Set parameters
r1 <- 0.5  # Intrinsic growth rate of species 1
r2 <- 0.7  # Intrinsic growth rate of species 2
alpha1 <- 0.02  # Intraspecific competition coefficient for species 1
alpha2 <- 0.03  # Intraspecific competition coefficient for species 2
beta1 <- 0.01  # Interspecific competition coefficient for species 1
beta2 <- 0.02  # Interspecific competition coefficient for species 2

# Time vector
time <- seq(0, 100, by = 0.1)

# Initial population sizes
N1_0 <- 10
N2_0 <- 10

# Run simulations
library(deSolve)
LV_simulation <- ode(y = c(N1 = N1_0, N2 = N2_0), times = time, func = LV_equations, parms = c(r1 = r1, r2 = r2, alpha1 = alpha1, alpha2 = alpha2, beta1 = beta1, beta2 = beta2))

# Create phase plane diagram
ggplot(data = as.data.frame(LV_simulation), aes(x = N1, y = N2)) +
  geom_path() +
  geom_point(aes(color = "Trajectory")) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "blue") +
  geom_abline(intercept = r1/alpha1, slope = -beta1/alpha1, linetype = "solid", color = "darkred") +
  geom_abline(intercept = r2/alpha2, slope = -beta2/alpha2, linetype = "solid", color = "forestgreen") +
  labs(x = "Species 1", y = "Species 2", color = "Legend") +
  theme_minimal() +
  scale_color_manual(values = c("Trajectory" = "black")) +
  ggtitle("Phase Plane Diagram: Lotka-Volterra Model") +
  theme(plot.title = element_text(hjust = 0.5))
```

### Reflective Questions: 
* What do the zero net growth isoclines represent in this diagram?

* How do the trajectories of species populations change over time? 
* What does this indicate about their interactions?

* How might changes in parameter values alter the stability of the system depicted in the phase plane diagram?

* Consider real-world ecosystems. Can you think of situations where the assumptions of the Lotka-Volterra model might not hold true?

* How would you extend this model to include more than two competing species? 
* What challenges might arise in such a scenario?

## This could be a different post.


### Monod's equation

Recall that the Monod models is a common model to predict microorganism growth.  


```{r}
# Function to simulate Monod model for planktonic bacteria
monod_model <- function(X, mu_max, Ks, S) {
  mu = mu_max * S / (Ks + S)
  dXdt = mu * X
  return(dXdt)
}

# Simulation parameters
time <- seq(0, 10, by = 0.1)  # Time points
S0 <- 10  # Initial substrate concentration
X1_0 <- 1  # Initial biomass concentration of species 1
X2_0 <- 1  # Initial biomass concentration of species 2
mu_max1 <- 1  # Maximum specific growth rate of species 1
mu_max2 <- 1.1  # Maximum specific growth rate of species 2 (slightly higher for competitive advantage)
Ks <- 2  # Half-saturation constant (equal for both species)

# Simulate Monod model for species 1
output1 <- data.frame(time = time)
output1$Biomass <- NA
output1$Biomass[1] <- X1_0
for (i in 2:length(time)) {
  S <- S0
  dXdt <- monod_model(output1$Biomass[i-1], mu_max1, Ks, S)
  output1$Biomass[i] <- output1$Biomass[i-1] + dXdt * (time[i] - time[i-1])
}

# Simulate Monod model for species 2
output2 <- data.frame(time = time)
output2$Biomass <- NA
output2$Biomass[1] <- X2_0
for (i in 2:length(time)) {
  S <- S0
  dXdt <- monod_model(output2$Biomass[i-1], mu_max2, Ks, S)
  output2$Biomass[i] <- output2$Biomass[i-1] + dXdt * (time[i] - time[i-1])
}

# Plot biomass dynamics for both species
ggplot() +
  geom_line(data = output1, aes(x = time, y = Biomass, color = "Species 1")) +
  geom_line(data = output2, aes(x = time, y = Biomass, color = "Species 2")) +
  labs(x = "Time", y = "Biomass", color = "Species") +
  theme_minimal()
```

### Contois model of competition

The Contois model is a classic model used to describe microbial growth in a chemostat environment, where bacteria are competing for limited resources. In this model, the growth rate of a species is proportional to the resource uptake rate and inversely proportional to the concentration of the limiting resource.

```{r echo=TRUE}
# Function to simulate Contois model
contois_model <- function(X, mu_max, Ks, Yxs, S) {
  # X: Biomass concentration
  # mu_max: Maximum specific growth rate
  # Ks: Half-saturation constant
  # Yxs: Yield coefficient
  # S: Substrate concentration
  
  mu = mu_max * S / (Ks + S)
  dXdt = mu * X
  return(dXdt)
}
```

Now lets set some simulation parameters

```{r}

time <- seq(0, 100, by = 0.1)  # Time points
S0 <- 10  # Initial substrate concentration
X1_0 <- 1  # Initial biomass concentration of species 1
X2_0 <- 0.5  # Initial biomass concentration of species 2
mu_max1 <- 1  # Maximum specific growth rate of species 1
mu_max2 <- 1.2  # Maximum specific growth rate of species 2
Ks1 <- 2  # Half-saturation constant of species 1
Ks2 <- 3  # Half-saturation constant of species 2
Yxs1 <- 0.5  # Yield coefficient of species 1
Yxs2 <- 0.6  # Yield coefficient of species 2
```

```{r}
# Simulate Contois model for species 1
output1 <- data.frame(time = time)
output1$Species <- rep("Species 1", length(time))
output1$Biomass <- NA
output1$Biomass[1] <- X1_0
for (i in 2:length(time)) {
  S <- S0 - output1$Biomass[i-1] * Yxs1
  dXdt <- contois_model(output1$Biomass[i-1], mu_max1, Ks1, Yxs1, S)
  output1$Biomass[i] <- output1$Biomass[i-1] + dXdt * (time[i] - time[i-1])
}

# Simulate Contois model for species 2
output2 <- data.frame(time = time)
output2$Species <- rep("Species 2", length(time))
output2$Biomass <- NA
output2$Biomass[1] <- X2_0
for (i in 2:length(time)) {
  S <- S0 - (output1$Biomass[i-1] * Yxs1 + output2$Biomass[i-1] * Yxs2)
  dXdt <- contois_model(output2$Biomass[i-1], mu_max2, Ks2, Yxs2, S)
  output2$Biomass[i] <- output2$Biomass[i-1] + dXdt * (time[i] - time[i-1])
}

# Combine outputs
output <- rbind(output1, output2)

# Plot biomass dynamics
ggplot(output, aes(x = time, y = Biomass, color = Species)) +
  geom_line() +
  labs(x = "Time", y = "Biomass", color = "Species") +
  theme_minimal()
```
This plot above simulates the biomass dynamics for two bacterial species (Species 1 and Species 2) competing for a single limiting resource over time under the Contois model.

```{r}
# Function to simulate Monod model for planktonic bacteria
monod_model <- function(X, mu_max, Ks, S) {
  mu = mu_max * S / (Ks + S)
  dXdt = mu * X
  return(dXdt)
}

# Function to simulate Contois model for aggregating bacteria
contois_model <- function(X, mu_max, Ks, Yxs, S) {
  mu = mu_max * S / (Ks + S)
  dXdt = mu * X
  return(dXdt)
}

# Simulation parameters
time <- seq(0, 100, by = 0.1)  # Time points
S0 <- 10  # Initial substrate concentration
X_p <- 1  # Initial biomass concentration of planktonic bacteria
X_a <- 1  # Initial biomass concentration of aggregating bacteria
mu_max_p <- 1  # Maximum specific growth rate of planktonic bacteria
mu_max_a <- 1  # Maximum specific growth rate of aggregating bacteria (equal to planktonic for coexistence)
Ks_p <- 2  # Half-saturation constant of planktonic bacteria
Ks_a <- 2  # Half-saturation constant of aggregating bacteria (equal to planktonic for coexistence)
Yxs_a <- 0.5  # Yield coefficient of aggregating bacteria (equal to planktonic for coexistence)

# Simulate Monod model for planktonic bacteria
output_p <- data.frame(time = time)
output_p$Biomass <- NA
output_p$Biomass[1] <- X_p
for (i in 2:length(time)) {
  S <- S0 - output_p$Biomass[i-1]
  dXdt <- monod_model(output_p$Biomass[i-1], mu_max_p, Ks_p, S)
  output_p$Biomass[i] <- output_p$Biomass[i-1] + dXdt * (time[i] - time[i-1])
}

# Simulate Contois model for aggregating bacteria
output_a <- data.frame(time = time)
output_a$Biomass <- NA
output_a$Biomass[1] <- X_a
for (i in 2:length(time)) {
  S <- S0 - output_a$Biomass[i-1] * Yxs_a
  dXdt <- contois_model(output_a$Biomass[i-1], mu_max_a, Ks_a, Yxs_a, S)
  output_a$Biomass[i] <- output_a$Biomass[i-1] + dXdt * (time[i] - time[i-1])
}

# Plot biomass dynamics
ggplot() +
  geom_line(data = output_p, aes(x = time, y = Biomass, color = "Planktonic Bacteria")) +
  geom_line(data = output_a, aes(x = time, y = Biomass, color = "Aggregating Bacteria")) +
  labs(x = "Time", y = "Biomass", color = "Species") +
	ggtitle("Coexistence of two bacterial types") +
  theme_minimal()
```
