---
title: "Competition"
description: |
  Competition with Lotka-Volterra models
author: Nathan Jud
date: 2024-05-13
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

### Learning Objectives:
1. Understand the basics of constructing a phase plane diagram for a Lotka-Volterra style model.
2. Differentiate between intraspecific and interspecific competition in ecological systems.
3. Identify zero net growth isoclines and their significance in population dynamics.
4. Explore the concept of niche and its relationship with the assumptions of the Lotka-Volterra model.
5. Introduce multi-species coexistence models as an extension of Lotka-Volterra.

### Preparation Questions:
1. What are the key differences between intraspecific and interspecific competition?
2. How do zero net growth isoclines help us understand the stability of population dynamics?
3. In what ways do the assumptions of the Lotka-Volterra model simplify real-world ecological systems?
4. Can you think of examples where species coexist in an ecosystem despite competing for the same resources?
5. How might adding more species to the Lotka-Volterra model affect its complexity and predictive power?

```{r echo=FALSE}
# Load required libraries
library(ggplot2)
```

```{r}
# Define Lotka-Volterra equations
LV_equations <- function(t, N, params) {
  with(as.list(c(N, params)), {
    dN1 <- r1 * N1 * (1 - (N1 * alpha1 + N2 * beta1))
    dN2 <- r2 * N2 * (1 - (N1 * alpha2 + N2 * beta2))
    return(list(c(dN1, dN2)))
  })
}
```

```{r}
# Set parameters
r1 <- 0.5  # Intrinsic growth rate of species 1
r2 <- 0.7  # Intrinsic growth rate of species 2
alpha1 <- 0.02  # Intraspecific competition coefficient for species 1
alpha2 <- 0.03  # Intraspecific competition coefficient for species 2
beta1 <- 0.01  # Interspecific competition coefficient for species 1
beta2 <- 0.02  # Interspecific competition coefficient for species 2

# Time vector
time <- seq(0, 100, by = 0.1)

# Initial population sizes
N1_0 <- 10
N2_0 <- 10

# Run simulations
library(deSolve)
LV_simulation <- ode(y = c(N1 = N1_0, N2 = N2_0), times = time, func = LV_equations, parms = c(r1 = r1, r2 = r2, alpha1 = alpha1, alpha2 = alpha2, beta1 = beta1, beta2 = beta2))

# Create phase plane diagram
ggplot(data = as.data.frame(LV_simulation), aes(x = N1, y = N2)) +
  geom_path() +
  geom_point(aes(color = "Trajectory")) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "blue") +
  geom_abline(intercept = r1/alpha1, slope = -beta1/alpha1, linetype = "solid", color = "darkred") +
  geom_abline(intercept = r2/alpha2, slope = -beta2/alpha2, linetype = "solid", color = "forestgreen") +
  labs(x = "Species 1", y = "Species 2", color = "Legend") +
  theme_minimal() +
  scale_color_manual(values = c("Trajectory" = "black")) +
  ggtitle("Phase Plane Diagram: Lotka-Volterra Model") +
  theme(plot.title = element_text(hjust = 0.5))
```

### Reflective Questions: 
* What do the zero net growth isoclines represent in this diagram?

* How do the trajectories of species populations change over time? 
* What does this indicate about their interactions?

* How might changes in parameter values alter the stability of the system depicted in the phase plane diagram?

* Consider real-world ecosystems. Can you think of situations where the assumptions of the Lotka-Volterra model might not hold true?

* How would you extend this model to include more than two competing species? 
* What challenges might arise in such a scenario?
